Associations between Chlorophyll \emph{a} and various Microcystin-LR Health Advisory Concentrations

### 2015-09-10 manuscript/manuscript.Rmd
re-render drafty intro

<!-- Abstract is being wrapped in latex here so that all analysis can be run in the chunk above and the results reproducibly referenced in the abstract. -->
Cyanobacteria harmful algal blooms (cHABs) are associated with a wide range of adverse health effects that stem mostly from the presence of cyanotoxins.  To help protect against these impacts, several health advisory levels have been set for some toxins.  In particular, one of the more common toxins, microcystin, has several advisory levels set for drinking water and recreational use and managing water bodies to meet those levels could have far reaching benefits.  However, compared to other water quality measures, measurements of microcystin are not common and current field measurement techniques have limited precision and accuracy.  Addressing these issues will take time and resourced.  Thus, there is utility in finding indicators of microcystin that are already widely available, can be estimated quickly and *in situ*, and used as a first defense against high levels of microcystin.  In particular, chlorophyll *a* is very commonly measured, can be estimated *in situ*, and has been shown to be positively associated with microcystin.  In this paper we use this association to provide estimates of chlorophyll *a* that if exceeded would be indicative of a higher probability of exceeding select health advisory concentrations for microcystin-LR.  Using the 2007 National Lakes Assessment and a conditional probability approach that has been used in other water quality settings, we identify chlorophyll *a* concentrations that are more likely than not to be associated with an exceedance of a microcystin health advisory level.  We look at the recent US EPA standards for drinking water as well as the World Health Organization levels for drinking water and recreational use.  For microcystin concentrations of 0.3, 1, 1.6, and 2 we find chlorophyll *a* concentrations of `r mc_chla[1]`, `r mc_chla[2]`, `r mc_chla[3]`, and `r mc_chla[4]`, respectively.  When managing for these various microcystin levels exceeding these reported chlorophyll *a* concentrations should be a trigger for further testing and possible management action.

@@

One of these complications is that they rely on available measurements of microcystin-LR.  This toxin can be measured in the field using test strips but these are a coarse measure at best and currently available test strips focus on 1 and 10 $\mu$g/L [REFS].  Measurements with greater accuracy and precision require taking regular water samples and having those samples processed in a lab to determine the toxin concentration [REFS].  This has the potentially to be costly and time consuming both factors which could limit monitoring efforts.  Additionally, microcystin-LR is not currently a routine water quality measurment and thus availability of data will continue to be a problem. Fortunately, microcystin-LR has been shown to be associated with several other, more easily measured components of water quality.

Chlorophyll *a* is a very commonly measured components of water quality that is also known to be associated with Microsystin-LR concentrations [REFS].  Additionally there are many rapid measurements for assessing chlorophyll *a* levels *in situ*.  For instance, there are small or hand held flourometers that provide reliable measurements [REFS].   Given these facts, it might be possible to identify chlorophyll *a* concentrations that would be associated with the various Microcystin-LR health advisory levels.  Identifying these associations would provide another reliable tool for water resource managers to use to help manage the threat to public health posed by cHABs.

Use association and cpa to id chl a concentration that indicative of exceeding HA

>>>>

<!-- Abstract is being wrapped in latex here so that all analysis can be run in the chunk above and the results reproducibly referenced in the abstract. -->
Cyanobacteria harmful algal blooms (cHABs) are associated with a wide range of adverse health effects that stem mostly from the presence of cyanotoxins.  To help protect against these impacts, several health advisory levels have been set for some toxins.  In particular, one of the more common toxins, microcystin, has several advisory levels set for drinking water and recreational use and managing water bodies to meet those levels could have far reaching benefits.  However, compared to other water quality measures, measurements of microcystin are not common and current field measurement techniques have limited precision and accuracy.  Addressing these issues will take time and resources.  Thus, there is utility in finding indicators of microcystin that are already widely available, can be estimated quickly and *in situ*, and used as a first defense against high levels of microcystin.  In particular, chlorophyll *a* is very commonly measured, can be estimated *in situ*, and has been shown to be positively associated with microcystin.  In this paper we use this association to provide estimates of chlorophyll *a* that if exceeded would be indicative of a higher probability of exceeding select health advisory concentrations for microcystin-LR.  Using the 2007 National Lakes Assessment and a conditional probability approach that has been used in other water quality settings, we identify chlorophyll *a* concentrations that are more likely than not to be associated with an exceedance of a microcystin health advisory level.  We look at the recent US EPA standards for drinking water as well as the World Health Organization levels for drinking water and recreational use.  For microcystin concentrations of 0.3, 1, 1.6, and 2 we find chlorophyll *a* concentrations of `r mc_chla[1]`, `r mc_chla[2]`, `r mc_chla[3]`, and `r mc_chla[4]`, respectively.  When managing for these various microcystin levels exceeding these reported chlorophyll *a* concentrations should be a trigger for further testing and possible management action.

@@

One of these complications is that they rely on available measurements of microcystin-LR.  This toxin can be measured in the field using test strips but these are a coarse measure at best and currently available test strips focus on 1 and 10 $\mu$g/L [REFS].  Measurements with greater accuracy and precision require taking water samples and processing those in a lab to determine the toxin concentration [REFS].  Additionally, microcystin-LR is, currently, not a routinely collected water quality parameter,thus, availability of microcystin-LR data may limit our ability to screen water bodies for exceedances of the various health advisories concentrations.  Until microcystin-LR concentrations are more widely collected an alternative measure is needed. Fortunately, microcystin-LR has been shown to be associated with several other, more easily measured components of water quality.

Chlorophyll *a* is a very commonly measured components of water quality that is also known to be associated with Microsystin-LR concentrations [REFS].  Additionally there are many rapid measurements for assessing chlorophyll *a* levels *in situ*.  For instance, there are small or hand held flourometers that provide reliable measurements [REFS].   Given these facts, it might be possible to identify chlorophyll *a* concentrations that would be associated with the various Microcystin-LR health advisory levels.  Identifying these associations would provide another reliable tool for water resource managers to use to help manage the threat to public health posed by cHABs and would be especially useful in the absence of microcystin-LR concentrations.  Thus, the goal of this paper is to utlize the National Lakes Assessment data and identify chlorophyll *a* concentrations that are associated with higher probabilities of exceeding several microcystin-LR health advisory concentrations [NLA REF].  So that others may repeat this analysis, the data, code, and this manuscript are freely available via [https://github.com/USAPE/microcystinchla](https://github.com/USAPE/microcystinchla).

### 2015-09-10 manuscript/manuscript.tex
re-render drafty intro

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
range of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect against these impacts, several health
advisory levels have been set for some toxins. In particular, one of the
more common toxins, microcystin, has several advisory levels set for
drinking water and recreational use and managing water bodies to meet
those levels could have far reaching benefits. However, compared to
other water quality measures, measurements of microcystin are not common
and current field measurement techniques have limited precision and
accuracy. Addressing these issues will take time and resourced. Thus,
there is utility in finding indicators of microcystin that are already
widely available, can be estimated quickly and \emph{in situ}, and used
as a first defense against high levels of microcystin. In particular,
chlorophyll \emph{a} is very commonly measured, can be estimated
\emph{in situ}, and has been shown to be positively associated with
microcystin. In this paper we use this association to provide estimates
of chlorophyll \emph{a} that if exceeded would be indicative of a higher
probability of exceeding select health advisory concentrations for
microcystin-LR. Using the 2007 National Lakes Assessment and a
conditional probability approach that has been used in other water
quality settings, we identify chlorophyll \emph{a} concentrations that
are more likely than not to be associated with an exceedance of a
microcystin health advisory level. We look at the recent US EPA
standards for drinking water as well as the World Health Organization
levels for drinking water and recreational use. For microcystin
concentrations of 0.3, 1, 1.6, and 2 we find chlorophyll \emph{a}
concentrations of 25.1, 69.6, 84.96, and 113.14, respectively. When
managing for these various microcystin levels exceeding these reported
chlorophyll \emph{a} concentrations should be a trigger for further
testing and possible management action.

@@

One of these complications is that they rely on available measurements
of microcystin-LR. This toxin can be measured in the field using test
strips but these are a coarse measure at best and currently available
test strips focus on 1 and 10 \(\mu\)g/L {[}REFS{]}. Measurements with
greater accuracy and precision require taking regular water samples and
having those samples processed in a lab to determine the toxin
concentration {[}REFS{]}. This has the potentially to be costly and time
consuming both factors which could limit monitoring efforts.
Additionally, microcystin-LR is not currently a routine water quality
measurment and thus availability of data will continue to be a problem.
Fortunately, microcystin-LR has been shown to be associated with several
other, more easily measured components of water quality.

Chlorophyll \emph{a} is a very commonly measured components of water
quality that is also known to be associated with Microsystin-LR
concentrations {[}REFS{]}. Additionally there are many rapid
measurements for assessing chlorophyll \emph{a} levels \emph{in situ}.
For instance, there are small or hand held flourometers that provide
reliable measurements {[}REFS{]}. Given these facts, it might be
possible to identify chlorophyll \emph{a} concentrations that would be
associated with the various Microcystin-LR health advisory levels.
Identifying these associations would provide another reliable tool for
water resource managers to use to help manage the threat to public
health posed by cHABs.

Use association and cpa to id chl a concentration that indicative of
exceeding HA

>>>>

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
range of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect against these impacts, several health
advisory levels have been set for some toxins. In particular, one of the
more common toxins, microcystin, has several advisory levels set for
drinking water and recreational use and managing water bodies to meet
those levels could have far reaching benefits. However, compared to
other water quality measures, measurements of microcystin are not common
and current field measurement techniques have limited precision and
accuracy. Addressing these issues will take time and resources. Thus,
there is utility in finding indicators of microcystin that are already
widely available, can be estimated quickly and \emph{in situ}, and used
as a first defense against high levels of microcystin. In particular,
chlorophyll \emph{a} is very commonly measured, can be estimated
\emph{in situ}, and has been shown to be positively associated with
microcystin. In this paper we use this association to provide estimates
of chlorophyll \emph{a} that if exceeded would be indicative of a higher
probability of exceeding select health advisory concentrations for
microcystin-LR. Using the 2007 National Lakes Assessment and a
conditional probability approach that has been used in other water
quality settings, we identify chlorophyll \emph{a} concentrations that
are more likely than not to be associated with an exceedance of a
microcystin health advisory level. We look at the recent US EPA
standards for drinking water as well as the World Health Organization
levels for drinking water and recreational use. For microcystin
concentrations of 0.3, 1, 1.6, and 2 we find chlorophyll \emph{a}
concentrations of 25.1, 69.6, 84.96, and 113.14, respectively. When
managing for these various microcystin levels exceeding these reported
chlorophyll \emph{a} concentrations should be a trigger for further
testing and possible management action.

@@

One of these complications is that they rely on available measurements
of microcystin-LR. This toxin can be measured in the field using test
strips but these are a coarse measure at best and currently available
test strips focus on 1 and 10 \(\mu\)g/L {[}REFS{]}. Measurements with
greater accuracy and precision require taking water samples and
processing those in a lab to determine the toxin concentration
{[}REFS{]}. Additionally, microcystin-LR is, currently, not a routinely
collected water quality parameter,thus, availability of microcystin-LR
data may limit our ability to screen water bodies for exceedances of the
various health advisories concentrations. Until microcystin-LR
concentrations are more widely collected an alternative measure is
needed. Fortunately, microcystin-LR has been shown to be associated with
several other, more easily measured components of water quality.

Chlorophyll \emph{a} is a very commonly measured components of water
quality that is also known to be associated with Microsystin-LR
concentrations {[}REFS{]}. Additionally there are many rapid
measurements for assessing chlorophyll \emph{a} levels \emph{in situ}.
For instance, there are small or hand held flourometers that provide
reliable measurements {[}REFS{]}. Given these facts, it might be
possible to identify chlorophyll \emph{a} concentrations that would be
associated with the various Microcystin-LR health advisory levels.
Identifying these associations would provide another reliable tool for
water resource managers to use to help manage the threat to public
health posed by cHABs and would be especially useful in the absence of
microcystin-LR concentrations. Thus, the goal of this paper is to utlize
the National Lakes Assessment data and identify chlorophyll \emph{a}
concentrations that are associated with higher probabilities of
exceeding several microcystin-LR health advisory concentrations {[}NLA
REF{]}. So that others may repeat this analysis, the data, code, and
this manuscript are freely available via
\href{https://github.com/USAPE/microcystinchla}{\url{https://github.com/USAPE/microcystinchla}}.
